---
title: "Ggplot2 Tutorial"
author: "Reilly Orman"
date: "4/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ggplot Description

Data visualization is key in analyzing data and understanding relationships between aspects of your data. 

Ggplot is one of the main packages used to visualize data, which is part of the `tidyverse` package. 

Some of the aspects that I will be going over in this tutorial are the basic functions and requirements in ggplot, such as the data, aesthetic mappings (`aes()`), and geom functions. 

The functions I will be going over in this tutorial are:

1. `geom_point`
2. `geom_bar`
3. `geom_density`
4. `geom_jitter`

## The Basics

First, we need to load the essential packages for using ggplot2. You can either install/load `tidyverse` or `ggplot2`. 

I am also going to load the `gapminder` data set, which I am going to use to demonstrate some functions of ggplot2.

```{r}
#load required packages
library(tidyverse)
library(gapminder)
```


