---
title: "Ggplot2 Tutorial"
author: "Reilly Orman"
date: "4/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Ggplot Description**

Data visualization is key in analyzing data and understanding relationships between aspects of your data. 

Ggplot is one of the main packages used to visualize data, which is part of the `tidyverse` package. 

Some of the aspects that I will be going over in this tutorial are the basic functions and requirements in ggplot, such as the data, aesthetic mappings (`aes()`), and geom functions. 

The functions I will be going over in this tutorial are:

1. `geom_point`
2. `geom_bar`
3. `geom_histogram`
4. `geom_density`

## **The Basics**

First, we need to load the essential packages for using ggplot2. You can either install/load `tidyverse` or `ggplot2`. 

I am also going to load the `penguins` data set from the `palmerpenguins` package, which I will use to demonstrate some functions of ggplot2.

```{r}
#load required packages
library(tidyverse)
library(palmerpenguins)
library(viridis)
```

The basic structure of making a ggplot graph would be the `ggplot()` function, your data, aesthetic mappings, and the `geom_function` you want to use. 

Here is an example of this structure.

```{r}
#  Scatterplot illustrating relationship between penguins' body mass and bill length   
ggplot(data = penguins) +
  aes(x = body_mass_g, 
      y = bill_length_mm) +
  geom_point()
```

As featured in this example, the ggplot function (`ggplot()`) is the first step in creating any data visualization with `ggplot2`. The data set I used was the penguins data set, and I separated each step with a + sign, which signals that you're adding to your graph. 

## **Aesthetics**

Then, I set the aesthetics with the `aes()` mappings, which connects your variables with visual dimensions. These aesthetics can be customized by color, fill, shape, and size.

If I wanted to recreate the previous plot by seeing the relationship between body mass and bill length by species, I would add either a color or fill that connects to the variable `species`. 

```{r}
# Scatterplot showing body mass and bill length by species
ggplot(data = penguins) +
  aes(x = body_mass_g,
      y = bill_length_mm,
      color = species) +
  geom_point() +
  scale_color_viridis_d()
```

If I wanted the whole plot to be one color that is not subset by species, then I would specify the color outside of the visual aesthetics and inside my `geom_function`. 

```{r}
# Scatterplot showing body mass and bill length outlined in purple
ggplot(data = penguins) +
  aes(x = body_mass_g,
      y = bill_length_mm) +
  geom_point(color = "purple")
```

## **Geom Functions**

There are dozens upon dozens of geom functions that have specific aesthetics to follow. In this tutorial, we are only going to focus on some of the more common and useful ones in respect to data visualization. 

### **`geom_point`**

The `geom_point()` function adds points to your graph, creating a scatterplot. This function is often used when there are two continuous variables being measured. 

What if we wanted to see a basic trend line from the last graph? For trend lines, the function `geom_smooth()` is used, which has a method aesthetic for the type of trend line presented. 

For a straight trend line, the `method = "lm"` is used, and for a curved trend line, `method = "loess"` is used. 

```{r}
#Scatterplot with straight trend line
ggplot(data = penguins) +
  aes(x = body_mass_g,
      y = bill_length_mm) +
  geom_point() +
  geom_smooth(method = "lm")
  
```

```{r}
#Scatterplot with curved trend line
ggplot(data = penguins) +
  aes(x = body_mass_g,
      y = bill_length_mm) +
  geom_point() +
  geom_smooth(method = "loess")
```

## **`geom_bar`**

The `geom_bar()` function plots a categorical variable. 

The basic structure of geom_bar is outlined below. 

```{r eval = FALSE}
#geom_bar template
ggplot(data = <dataset>) +
  aes(x = <categorical variable>,
      y = <numerical variable>) +
  geom_bar(color = "named color",
           fill = "named color",
           stat = <statistical computation>)
```

If we wanted to see the total number of each species, we would use this simple structure of a bar chart outlined below. 

```{r}
#Bar chart for penguin species
ggplot(data = penguins) +
  aes(x = species) +
  geom_bar()
```

As defaulted, the statistical computation for a bar chart in ggplot2 is bin, where the number of penguins by species is plotted on the y axis. 

If you wanted to plot one variable on the x-axis and another on the y-axis, you would set `stat = "identity"`. This enables you to plot the length of the bars to another variable. For example, if we wanted to create a bar chart illustrating different species bill length, here's how we would do it. 

```{r}
#remove rows with missing values and assign that data set to a new object
new_penguins <- na.omit(penguins)

#subset data set by summarizing the mean bill length for each species, then creating bar chart plotting summary table
new_penguins |> 
  group_by(species) |> 
  summarize(mean = mean(bill_length_mm)) |> 
  ggplot() +
  aes(x = species,
      y = mean, fill = species) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d()
```

## **`geom_histogram`**

`geom_histogram()` is often used when plotting a continuous variable. 

Here is a basic example of the `geom_histogram()` function with the penguins dataset.

```{r}
#basic histogram for flipper length
ggplot(new_penguins) +
  aes(x = flipper_length_mm) +
  geom_histogram(bins = 10)
```

In this histogram, you are also able to compare the flipper length by species using the color or fill argument within aesthetics. 

```{r}
#histogram for flipper length by species
ggplot(new_penguins) +
  aes(x = flipper_length_mm, 
      fill = species) +
  geom_histogram(alpha = .5, position = "identity", color = "black", bins = 10) +
  scale_fill_viridis_d()
```

### **Facets**

Another way to look at flipper length by species would be to use facets, which splits the plot into panels. 

The way to use facets is either by `facet_grid()` or `facet_wrap`.

#### **`facet_grid()`**

`facet_grid()` plots the variables in a 2d grid, either by `. ~ <variable>` or `<variable> ~ .`.

* `. ~ <variable>`
  + spreads the values of the variable across columns, having the vertical scales aligned. 
* `<variable> ~ .`
  + spreads the values of the variable across rows, having the horizontal scales aligned. 

#### **`facet_wrap()`**

`facet_wrap()` creates the ribbon of panels that are generated by the levels of the variables. 

This function can be especially useful when there are many levels in one variable, and you are trying to save space on your graphs. 
  
Below are two histograms comparing facet grid to facet wrap.  

```{r}
#flipper length by species in histogram using facet grid
ggplot(new_penguins) +
  aes(x = flipper_length_mm) +
  geom_histogram(color = "black", bins = 10) +
  facet_grid(species ~ .)
```

```{r}
#flipper length by species in histogram using facet wrap
ggplot(new_penguins) +
  aes(x = flipper_length_mm) +
  geom_histogram(color = "black", bins = 10) +
  facet_wrap(~ species)
```

Based on these histograms, `facet_grid()` seems better suited to illustrate our data in a meaningful way. I am able to compare the flipper length by species more directly by aligning the horizontal scales. Since there are only three levels of the species variable, facet wrap is not needed to save space. 

### **Setting the bins**

I set the number of bins to 10 using Sturge's Rule. The formula is:

$$ K = 1 + 3.322 log_N $$ 
K = number of bins.

N = number of observations in data.

## **`geom_density`**

The `geom_density()` function plots the distribution of a numeric variable. 

Here is a basic example of using the `geom_density()` function with the penguins data set. 

```{r}
#basic density plot for flipper length
ggplot(new_penguins) +
  aes(x = flipper_length_mm) +
  geom_density()
```

### **Using density plot with histograms**

You're able to combine multiple plots together on the same scale, such as density and histogram plots. To do this, plot the histogram with density on the y-axis instead of count. 

```{r}
#density and histogram plot
ggplot(new_penguins) +
  aes(x = flipper_length_mm) +
  geom_histogram(aes(y = ..density..), color = "black", bins = 10) +
  geom_density(alpha = .5, fill = "blue")
```

